---
title: "STAT545 HW-04"
author: "Ernest Lai"
output: md_document
---

## R Markdown

This is the R Markdown document for hw-04, which is a data wrangling exercise of the **gapminder** dataset. The focus of this assignment would be on data reshaping (and relationship to aggregation) and data join. 

For simple reshaping, gather() and spread() from [*tidyr*](https://tidyr.tidyverse.org/) - to help a more tidier dataset. For data joining, there are two data sources and info from both datasets are needed - hence a multitude of join prompts will be used to  c ombine info from both datasets into a single new object.

All of the functions used for the data reshaping and data join will be available under the [*tidyverse*](https://www.tidyverse.org/packages/) package. For full information about hw-04, please visit [here](http://stat545.com/Classroom/assignments/hw04/hw04.html).

<br/>

## Open the tidyverse and gapminder Package

As usual, we will begin with loading the packages needed for the analysis (tidyverse for functions and gapminder as the dataset of interest). If these packages are being used for the first time on your R-Studio client, then the packages will need to be installed first prior to being loaded.

```{r open packages, warning = FALSE, message = FALSE}
# If loading tidyverse and gapminder for first time, then run the install.packages function below before loading the library.

# install.packages('tidyverse')
# install.packages('gapminder')

# Load tidyverse and gapminder
library(tidyverse)
library(gapminder)
```
<br/>

## Data Reshaping Prompt

For the data reshaping exercise, I will complete **Activity 2: Making a tibble with one row per year and columns for life expectancy for two or more countries**. The newly reshaped data frame will be put in table form via **knitr::kable()**. Lastly, a scatterplot comparing life expectancies from one country to another will be generated from the new reshaped data set.

<br/>

### Reshaping the Data + Tabular Presentation

We will first assign the regular gapminder dataset as **gm_untidy**, work on reshaping this object to show life expectancies by year for 5 countries (1 for each geographic area according to the gapminder dataset). We will keep year, lifeExp, and country as these are the variables of interest we would wish to focus on. 

For this activity, there were two main choices with how we wanted to reshape the data: *gather()* or *spread()*. In this scenario, we are taking a specific variable (country) and trying to present life expectancy at each timepoint/every 5 years. Since we are using a categorical variable and creating new columns based on the countries we are interested in, we have to **spread** the data. 

In this scenario, we are trying to take the untidy data and create  wider data frame - hence we take two important columns: country, which is our key variable to expand on, and life expectancy, our value which we want to present for each country of interest for all timepoints. In other words, we are "spreading" each unique country in our country variable into their own columns and presenting the life expectancy value for each of their timepoint.

The gather function will not work in this scenario because it takes multiple columns from the untidy data and puts them into key-value pairs, making our new data frame longer rather than stretching the data frame.

After reshaping our data using spread(), the results will be published on a table.

```{r, data-reshaping-table, warning = FALSE, message = FALSE}
# Set gapminder as a data frame object
gm_untidy <- gapminder

# Set the data frame as a tibble (optional, as gm_untidy.1 is already a data frame). It is good practice to use tibble to create a data frame as it retains the input type for each variable.
tibble <- as_tibble(gm_untidy)

# Pipe tibble through df modification and reshaping functions, then present the end result in a table (see comments for explanation of each function). Store new reshaped data frame as gm_tidy, for plotting purposes later.
gm_tidy <- tibble %>%
  
# Capitalize year in our dataset, to be used in table for presenting reshaped data  
    rename(Year = year) %>%
  
# select variables of interest from gm_untidy.1
    select("country", "Year", "lifeExp") %>% 
  
# select countries of interest from the gapminder dataset
    filter(country %in% c("Canada", "France", "Japan", "South Africa", "New Zealand")) %>%
  
# take the country (key) column, spread all unique countries into its own column, then populate life expectancy (value) with respect to the specified country  
    spread(key = "country", value = "lifeExp") 

# present newly reshaped data frame into a table, with additional formatting (preserve row and column labels)
gm_tidy %>%
knitr::kable(caption = "This table summarizes the life expectancy every 4 years for Canada, France, Japan, New Zealand, and South Africa")
```



For the most part, the 5 countries of interest in our data analysis show improvement in life expectancy at each subsequent timepoint/every 5 years. The only exception is SOuth Africa, where there is a small decline from 1992 to 1997, but the saw sharper decline for the 2 subsequent timepoints after 1997. Life expectancy is generally in the same range between Canada, Japan, France, and New Zealand, but it is noticeable lower comparatively for South Africa.

The next section will show a scatterplot of life expectancy across all timepoints for the 5 countries. This will allow to make life expectancy comparisons between countries based on graphical observations.

<br/>

### Life Expectancy Plots

Two plots will be presented: An individual scatter plot of life expectancy across all timepoints for each country from the untidy data set, and a scatterplot comparing Canada's life expectancy to South Africa and looking at their trend/relationship.

```{r, untidy comparison scatter}
# Start a scatterplot with the untidy data first
tibble %>%
  
# select variables of interest from gm_untidy.1
    select("country", "year", "lifeExp") %>% 
  
# select countries of interest from the gapminder dataset
    filter(country %in% c("Canada", "France", "Japan", "South Africa", "New Zealand")) %>%

# Plot life expectancies of all countries on one scatterplot, colour code points by country
  ggplot(aes(year, lifeExp), group = country) + # plot life expectancy against year
  geom_point(aes(colour = country), alpha = 1) + # lifeExp for each country every 5 years
  geom_line(aes(colour = country)) + # plot line through points, colour coded by country

# Set axis limits, add title (main and axis), and descriptions
  ylim(40, 85) + # limit for y-axis
  labs(title = "Comparison of Life Expectancy between all 5 countries", 
       y = "Life Expectancy (in Years)", x ="Year",
       caption = "Figure 1. Life expectancy every 5 years between South Africa and Canada",
       colour = "Country") + 
  theme_bw() # give a white background to the graph

```
<br/>

The visual plot from the untidy dataset confirm our observations from the tabular data. Canada, France, Japan, and New Zealand have similar life expectancies across all timepoints and seem to trend towards and improvement with every subsequent timepoint. Japan did have a lower life expectancy than the three other countries to start, but quickly caught up in 1967. South Africa in comparison has noticeably lower life expectancy than Canada, France, Japan, and New Zealand. The life expectancy for South Africa improves with each subsequent timepoint, but saw a sharp decline from 1997 to 2007.

For the next plot, we will use the reshaped data set to plot life expectancies of South Africa against the life expectancies of Canada across all time points. I chose Canada because I have lived here for almost my whole life, and I am comparing it to South Africa because this country had a noticeably different tracjectory in life expectancy across all timepoints compared to Japan, New Zealand and France.

```{r, tidy comparison scatter}
# Create scatter plot, with Canada and Africa on the x- and y-axis respectively. 
gm_tidy %>%

# Convert Year to factor for this exercise, to give contrasting colors for each data point
  mutate(Year = factor(Year, levels = c(1952, 1957, 1962, 1967, 1972, 1977, 1982, 1987, 1992,
                                        1997, 2002, 2007))) %>%
  
# Rename South Africa (empty space not allowed when selecting variables for analysis)
  rename(South_Africa = "South Africa") %>%
  
# Plot life expectancies of South Africa against Canada
  ggplot(aes(Canada, South_Africa), group = Year) + # Plot Canada against South Africa
  geom_point(aes(colour = Year), alpha = 1) + # plot lifeExp at each year/timepoint
  geom_line(alpha = 0.2) + # plot line through all points in order of timepoint

# Set axis limits, add title (main and axis), and descriptions
  xlim(65, 85) + # limit for x-axis
  ylim(45, 65) + # limit for y-axis
  labs(title = "Comparison of Life Expectancy between South Africa and Canada", 
       y = "South Africa", x ="Canada",
       caption = "Figure 2. Life expectancy every 5 years between South Africa and Canada") + 
  theme_bw() # give a white background to the graph
```
<br/>

This plot is a bit different from the first plot, as this plot allows the comparison between the rate of change in life expectancy between two countries (South Africa and Canada in this example). Even though South Africa has a noticeably lower life expectancy at the start compared to Canada, the life expectancies of South Africa at subsequent timepoints are improving at a faster rate than Canada - as evident by the steep positive slope. 

One would expect South Africa to eventually catch up to the other countries in a couple of decades if the trend continued, but the sharp decrease as mentioned earlier from  1997 to 2007 changed the course of the line plot. During this time period, the slope is negative because Canada's life expectancy was still improving at a steady slow rate but South Africa saw their life expectancy worsen in comparison to previous timepoints.

## Join Prompts

For this exercise, I will focus on creating a cheatsheet patterned after [Jenny's tutorial](http://stat545.com/bit001_dplyr-cheatsheet.html) on join prompts. My specific topics on this will be on soccer players, their nationality, the club they play for and the league name + country of the club they are in. 

### Load .csv Files Regarding Player and Team Data

There will be two spreadsheets for this section: One for player data and one for team data. The player data will consist of player name, nationality and current club. The team data will contain details on clubs, such as the league they are in and its nationality. Lets load the data:

```{r, load .csv, warning = FALSE, message = FALSE}
# Load tidyverse library if you haven't already
# library(tidyverse)

# Load player data and team data
library(readr) # load library for read_csv function

pdata <- read_csv("player_data.csv", skip = 0) # read player data, don't skip any rows
tdata <- read_csv("team_data.csv", skip = 0) # read team data, don't skip any rows

```

We loaded the tidyverse package earlier, the tidyverse package contains **dplyr** which provides the join functions we will be using. The **readr** package allows us to use the *read_csv()* function to upload the csv into our R-Studio environment.

### The Data

Working with two dataframes **pdata (x)** and **tdata (y)**, we will use *left_join*, *right_join*, *inner_join*, *semi_join*, *full_join*, and *anti_join*. The player data consists of players playing all around the world, from clubs in all continents. The club data consists of teams whom are very famous and globally known.

### left_join(x,y)
```{r, left_join}
left_join(pdata, tdata, by = "Club") %>% # join by variable 'Club'
knitr::kable(caption = "Left join between pdata (x) and tdata (y)")
```

The **left_join** function is a mutating join that returns all rows from 'x' (pdata) and adds columns from 'y' (tdata). This function matches pdata and tdata by the variable **Club**. If **Club** from pdata and tdata match then the function will bind additional information from tdata such as **League** and **League_Country** to pdata. If there are multiple matches between pdata and tdata by **Club**, all combination of the matches are returned.The new table tells us the club each player plays for, and the league and country the club is under.

We could also run left_join function without specifying which variable to match by, R is smart enough to fan out the identical column names between pdata and tdata and match on those columns. For future join functions, we will exclude the 'by' portion of the code - as we will be using pdata and tdata only.

### right_join(x,y)
```{r, right_join}
right_join(pdata, tdata) %>% # join by variable 'Club'
knitr::kable(caption = "Right join between pdata (x) and tdata (y)")
```

The **right_join** function is a mutating join that returns all rows from 'y' (pdata) and adds columns from 'x' (tdata). This function matches pdata and tdata by the variable **Club**. If **Club** from pdata and tdata match then the function will bind additional information from pdata such as **First_Name** and **Last_Name** to pdata. The new table tells us all club names, and adds the players from pdata playing for clubs that matched the list in tdata. 

You will notice there are new rows created, that is because there may be more than one player from pdata playing for the club listed from tdata. The NA correspond to the fact that no player from pdata has a **Club** matching the list of clubs under tdata.

### inner_join(x,y)
```{r, inner_join}
inner_join(pdata, tdata) %>% 
knitr::kable(caption = "Inner join between pdata (x) and tdata (y)")
```

The inner_join function returns all rows from pdata where there are matching values in tdata by the variable **Club**, and all columns from pdata and tdata. Since there are multiple matches by **Club**, as there are more than one player that plays for the same club under pdata, all combination of the matches are returned. This is also a mutating join, and we retained all rows from pdata because there aren't any clubs in pdata that are missing in tdata.

### semi_join(x,y)
```{r, semi_join}
semi_join(pdata, tdata) %>% 
knitr::kable(caption = "Semi join between pdata (x) and tdata (y)")
```

The semi_join function returns all rows from pdata where there are matching values in **Club** with tdata, but the semi_join keeps just the columns from x. Once again, all data is retained because all the entries in **Club** from pdata also exist in tdata.

### full_join(x,y)
```{r, full_join}
full_join(pdata, tdata) %>% 
knitr::kable(caption = "Full join between pdata (x) and tdata (y)")
```

The full_join function return all possible combinations (all rows and all columns from both pdata and tdata). When there are no matching values for **Club**, the table returns NA for whichever variable has missing entries.

### anti_join(x,y)
```{r, anti_join}
anti_join(pdata, tdata) # no table generated
```

Anti_join returns all rows from pdata where there are **non-matching** values for **Club** in tdata, and keeps only just the columns from pdata. This output returns no rows, because all unique values in **Club** from pdata matches the values of **Club** from tdata.

### Changing Arguments for Join Functions

It would be interesting to see how the join functions work, if we flip tdata to be our first argument and pdata as our second. The following set of join functions are a repeat of what I just went through, with the exception of the changed ordering of the datasets in the argument for each join function.

To spare you the details for each table, I will include the first 20 entries of the new table from the join functions

### left_join(y,x)
```{r, left_join_2}
lj<- left_join(tdata, pdata) 
knitr::kable(lj[1:20,], caption = "Left join between tdata (x) and pdata (y)")
```

The **left_join** function is a mutating join that returns all rows from tdata and adds columns from pdata. This function matches tdata and odata by the variable **Club**. If **Club** from tdata and pdata match then the function will bind additional information from pdata such as **First_Name**, **Last_Name**, and **National_Team** to tdata. For **Club** not matched between tdata and pdata,  there will be NA's from the new columns binded from pdata


### right_join(y,x)
```{r, right_join_2}
rj <- right_join(tdata, pdata) 
knitr::kable(rj[1:20,], caption = "Right join between tdata (x) and odata (y)")
```

The **right_join** function is a mutating join that returns all rows from 'x' (tdata) and adds columns from 'y' (pdata). This function matches tdata and pdata by the variable **Club**. If **Club** from tdata and pdata match then the function will bind additional information from pdata such as **First_Name**, **Last_Name**,and **Nationality** to tdata. The new table tells us all player names, and adds the team information from tdata, for clubs that matched the list in pdata. 


### inner_join(y,x)
```{r, inner_join_2}
ij <- inner_join(tdata, pdata)
knitr::kable(ij[1:20,], caption = "Inner join between tdata (x) and odata (y)")
```

The inner_join function returns all rows from tdata where there are matching values in pdata by the variable **Club**, and all columns from pdata and tdata. Since there are multiple matches by **Club**, as there are more than one player that plays for the same club under pdata, all combination of the matches are returned. This is also a mutating join, but the exception here is that we did not retain the rows from tdata for clubs that did not match the list in pdata.

### semi_join(y,x)
```{r, semi_join_2}
sj <- semi_join(tdata, pdata)
knitr::kable(sj[1:20,], caption = "Semi join between tdata (x) and pdata (y)")
```

The semi_join function returns all rows from tdata where there are matching values in **Club** with pdata, but the semi_join keeps just the columns from x. Once again, the exception here is that not all data from tdata is retained because there are entries in **Club** from tdata that do not exist in pdata.

### full_join(y,x)
```{r, full_join_2}
fj <- full_join(tdata, pdata)
knitr::kable(fj[1:20,], caption = "Full join between tdata (x) and pdata (y)")
```

The full_join function return all possible combinations (all rows and all columns from both tdata and pdata). When there are no matching values for **Club** between tdata and pdata, the table returns NA for whichever variable has missing entries from pdata. 

### anti_join(y,x)
```{r, anti_join_2}
aj <- anti_join(tdata, pdata)
knitr::kable(aj[1:17,], caption = "Anti join between tdata (x) and pdata (y)")
```

Anti_join returns all rows from tdata where there are **non-matching** values for **Club** in pdata, and keeps only just the columns from pdata. This output returned 17 rows, because there are 17 clubs from tdata which do not exist in the Club variable from tdata.

### Argument Order for Join Functions

Overall, order matters when using the join functions. As evidenced by some of the join prompts above, the ordering of datasets in the argument field can create differing output and also affects the ordering of the variables from the output